---
title: "PA1_template"
author: "Ravi Shankar"
date: "Sunday, May 17, 2015"
output: html_document
---
### set up caching and knitr chunk dependency calculation
``` {r setoptions,tidy=F, cache=F, eval=T, echo=F, results = "hide"}opts_chunk$set(autodep=T)dep_auto()
```

### Run the following data prep steps. Assumed that the working directory has a folder with .csv data:

### Part 1: Loading and preprocessing the data

```{r dataprep,tidy=F,cache=T}
dir<- "C:/Users/shankar/Desktop/Reproducible_Research/Data/PA1_submission"
setwd(dir)
dir <- getwd()
sub.dir <- "repdata_data_activity/activity.csv"
data.dir <- paste(dir, sub.dir, sep="/")
fit.steps <- read.csv(data.dir)
paste('checked at',date())
```
### Part 2: What is the mean total number of steps taken per day?
```{r part2steps12,tidy=F,cache=T}
#Remove Nas
fit.steps.good <- na.omit(fit.steps)
summary (fit.steps.good$date)
date.factor <- factor(fit.steps.good$date)
# Assignment, Part 2 step 1: Find total number of steps taken per day
day.sum<-tapply(fit.steps.good$steps, date.factor, sum)
day.sum
# Assignment, Part 2 step 2: Make a histogram of the total number of steps taken each day"
pdf("HistogramStepsNAIgnored.pdf")
hist(day.sum,breaks=100,col='red', xlab="Total Steps Per Day", main='Histogram')
dev.off()
```
##### The histogram was saved to HistogramStepsNAIgnored.pdf

```{r part2step 3,tidy=F,cache=T}
# Assignment, Part 2 step 3: Calculate and report mean and median of the total number of stepts taken per day
day.mean <-tapply(fit.steps.good$steps, date.factor, mean)
day.mean
day.median <-tapply(fit.steps.good$steps, date.factor, median)
day.median
```

##### The mean of the total number of steps taken per day are `r day.mean`. See above for a better formatted output along with dates.

##### The median of the total number of steps taken per day are `r day.median`.See above for a better formatted output along with dates.

### Part 3: What is the average daily activity pattern?

```{r part3step1,tidy=F,cache=T}
#Assignment Part 3 step 1: Time series plot of the 5-minute interval (x-axis)
# and the average number of steps taken, averaged across all days (y-axis)

interval.factor.good <- factor(fit.steps.good$interval)
interval.mean.good <- tapply(fit.steps.good$steps, interval.factor.good, mean)

# Make a time-series plot
pdf("TimeSeriesAvgStepsNAIgnored.pdf")
opar<-par(no.readonly=TRUE)
par(lty=1, pch=17)
plot(fit.steps.good$interval[1:288], interval.mean.good, type = "l", main = "Time Series Plot",ylab ="Average No. of Steps" ,xlab="5 min time intervals")
par(opar)
dev.off()
```
##### The Time Series Plot was saved to TimeSeriesAvgStepsNAIgnored.pdf

```{r part3step2,tidy=F,cache=T}
#Assignment Part 3 step 2: Find and report the max. number of steps
steps.max.good = max(interval.mean.good)
a<- interval.mean.good
b<- which(a==max(a))
time.at.max <- fit.steps.good$interval[b]
time.at.max
steps.max.good
```
##### The maximum number of steps is `r steps.max.good` at this 5-minute interval: `r time.at.max`.

### Part 4: Imputing missing values

```{r part4step1,tidy=F,cache=T}
# Part 4, step 1: Calculate and report the total # of missing values in the dataset
missing.flag <- is.na(fit.steps$steps)
no.missing <- table(missing.flag)["TRUE"]
no.missing 
```

##### The number of missing values in the dataset are `r no.missing`.

```{r part4step2,tidy=F,cache=T}
#Part 4, part 2: Devise a strategy to replace NAs in the dataset
# Replaced the missing values with the mean for the 5 minute interval

# First, tried replacing with the day average. This was not meaningful - day long data were missing
fit.steps.good <- na.omit(fit.steps)
# There were only 53 levels now.
date.factor.good <- factor(fit.steps.good$date)
day.mean.good <- tapply(fit.steps.good$steps, date.factor.good, mean)
# so, there were 8 days with no activity at all. So, this is not a good way
# to find means to plug into na slots.

# Second, tried to find the mean for a given 5 minute interval and may be there will be
# no na factors there (that is, it will return 288 both times)
# This was verified to be the case.
# So, mean for 5 minute intervals are calculated and are used in place of NAs 
interval.factor <- factor(fit.steps$interval)
interval.factor.good <- factor(fit.steps.good$interval)
interval.mean.good <- tapply(fit.steps.good$steps, interval.factor.good, mean)
```
##### Strategy used: Use the mean for 5-minute intervals to substitite for missing values. 

```{r part4step3,tidy=F,cache=T}
#Part 4, step 3: create a new dataset 
fit.index <- !is.na(fit.steps$steps)
# if fit.index is FALSE, change the NA to mean value for that interval
fs <- fit.steps
img <- interval.mean.good
# add img to fs data set. img is repeated for all the data sets
fs$mean <- img
# Just to compare and contrast, create another array - 'step' and update the
# NAs based on the index flag. All other values should remain the same. 
fs$step[!fit.index]<- fs$mean[!fit.index]
#Verified that the numbers are OK for the case of 3

new.fit.steps <- fit.steps
new.fit.steps$steps <- fs$step
```
##### Created a new dataset above that is equal to the orignial dataset but with the missing data filled in.

```{r part4step4ab,tidy=F,cache=T}
# Part 4, step 4: Make a histogram and calculate mean and median. Are they
# different?

summary (new.fit.steps$date)
new.date.factor <- factor(new.fit.steps$date)
# Assignment, Part 4, Step 4a: Find total number of steps taken per day
new.day.sum<-tapply(new.fit.steps$steps, new.date.factor, sum)
new.day.sum
# Assignment, Part 4, Step 4b: Make a histogram of the total number of steps taken each day"
pdf("HistogramStepsNAFilled.pdf")
hist(new.day.sum,breaks=100,col='red', xlab="Total Steps Per Day", main='Histogram')
dev.off()
```
```{r part4step4c,tidy=F,cache=T}
# assignment, Part 4, sub part 4: Calculate mean and median. Are they diffeent?
summary(new.fit.steps)
summary(fit.steps)
# No effect since the average is added to the NA and avearged. 
# As an example, With 10 values at 1, the average is 1. Now replace one NA with 1. 
#Take the new average for 11 values of 1 each. It is still 1. 
```
##### Total daily number of steps remains the same, as explained above in comments

### Part 5: Are there differences in activity patterns between weekdays and weekends?

```{r part5step1,tidy=F,cache=T}
# set up a vector of weekday (false) or weekend
#remove factor
date.char <- as.character(new.fit.steps$date)
step.date <- as.Date(date.char,'%Y-%m-%d')
# step.date
head(step.date)
tail(step.date)
day.name <- weekdays(step.date)
# day.name
head(day.name)
tail(day.name)
new.fit.steps$day.name <- day.name
# set up a flag -- to distinguish weekdays from weekend
wkend1 <- "Saturday"
wkend2 <- "Sunday"
weekend.flag <- (day.name == wkend1) | (day.name == wkend2)
head(weekend.flag)
tail(weekend.flag)
new.fit.steps$weekend.flag <- weekend.flag
#split based on weekday (FALSE) or weekend day (TRUE)

#calculate 5 minute interval means for weekdays and weekend days separately

new.fit.steps.weekday <- new.fit.steps[which (new.fit.steps$weekend.flag == "FALSE"),]
new.interval.weekday.factor  <- factor(new.fit.steps.weekday$interval)
new.fit.steps.weekend <- new.fit.steps[which (new.fit.steps$weekend.flag == "TRUE"),]
new.interval.weekend.factor  <- factor(new.fit.steps.weekend$interval)
new.interval.mean.weekday  <- tapply(new.fit.steps.weekday$steps, new.interval.weekday.factor, mean)
new.interval.mean.weekend  <- tapply(new.fit.steps.weekend$steps, new.interval.weekend.factor, mean)
```
##### Two sets of data for weekday and weekend day are obtained (see above)
#####Hello

```{r part5step2,tidy=F,cache=T}
#plot both weekday and weekend separately
opar<-par(no.readonly=TRUE)
par(lty=1, pch=17)
split.screen(c(2,1))
#set.panel(2,1)
pdf("TimeSeriesWeekdayWeekendStepsNAFilled.pdf")
screen(1)
plot(fit.steps.good$interval[1:288], new.interval.mean.weekday, type = "l", main = "Weekday Time Series Plot",ylab ="Average No. of Steps" ,xlab="5 min time intervals")
screen(2)
plot(fit.steps.good$interval[1:288], new.interval.mean.weekend, type = "l", main = "Weekend Time Series Plot",ylab ="Average No. of Steps" ,xlab="5 min time intervals")
dev.off()
#set.panel()
par(opar)
```
##### Two panel plots for weekday and weekend days is saved in TimeSeriesWeekdayWeekendStepsNAFilled.pdf (two plots). Much more activity during the whole weekend day(s) is noted. 

